---
title: "Heart Failure Analysis"
subtitle: "Handling Nonlinearity & Model Validation"
date: "23-04-2025"
author: "Ashley Dickson"
---

# Introduction
In this document, I build several models of the heart failure data that are able to handle non-linearities in the predictors, generate predictions and perform model validation.

# Setup
## Environment
Load the requisite libraries.

```{r, warning=FALSE, message=FALSE}
library(rms)
library(dplyr)
library(readxl)
library(tidyverse)
library(cardx)
library(gtsummary)
library(ggplot2)
library(GGally)
library(splines)
library(broom)
library(pROC)
library(caret)
library(corrplot)
library(CalibrationCurves)
library(knitr)
```

## Data
Having loaded the various libraries needed, let's import our heart failure data.

```{r}
setwd("C:/Users/rmhimdi/OneDrive - University College London/Documents/HeartFailure")
df <- read_excel("simulateddata_LR_NL.xlsx")
kable(head(df))
```


This is a Quarto website.
To learn more about Quarto websites visit <https://quarto.org/docs/websites>.